---
title: "hw03_figalit.Rmd"
output: github_document
---

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
library(ggplot2)
```

## Populating cheat sheet!

Get the maximum and minimum of GDP per capita for all continents.

```{r}
# One way is to present them stacked on top of each other.
minimums <- aggregate(gdpPercap ~ continent, gapminder, min)
maximums <- aggregate(gdpPercap ~ continent, gapminder, max)
combinedMinMaxData <- rbind(minimums, maximums)
knitr::kable(combinedMinMaxData)

# Or in a different way, "side by side":
minMaxData <- gapminder %>%
  group_by(continent) %>%
  summarize(minim = min(gdpPercap),
            maxim = max(gdpPercap))
knitr::kable(minMaxData)

# Let's plot out these points.
ggplot(combinedMinMaxData, aes(x = continent, y = gdpPercap)) +
  geom_point()

# And now all data in a boxplot
ggplot(gapminder, aes(x = continent, y = gdpPercap)) +
  geom_boxplot(outlier.colour = "blue") +
  geom_jitter(position = position_jitter(width = 0.1, height = 0), alpha = 1/4)
```

Look at the spread of GDP per capita within the continents.

Answer: To understand spread, we need to look at different measures, like standard deviation or variance, median or quantiles.
I have plotted some different ways of understanding this spread. 
```{r}
# Some statistics: 
data <- gapminder %>% 
  group_by(continent) %>%
  summarize(mean = mean(gdpPercap), 
            median = median(gdpPercap), 
            std = sd(gdpPercap), 
            interquantileRange = IQR(gdpPercap)) %>%
  arrange(std) # optional
knitr::kable(data)

# -- Some plots --
# One way is to look at a nice histogram to understand the general trend for each continent
ggplot(gapminder, aes(gdpPercap)) + 
  geom_histogram(aes(fill=continent), bins=15)

# And we could have many histograms, one per each continent
ggplot(gapminder, aes(gdpPercap)) + 
  geom_histogram(bins=15) + 
  facet_wrap( ~ continent)

# We could also have several plots, one per each year, 
# but Oceania does not have enough points for a violin plot.
ggplot(gapminder, aes(x=continent, y=gdpPercap, fill=continent)) +
  geom_violin() +
  scale_y_log10() +
  facet_wrap(~ year) +
  scale_fill_manual(values = c("blue", "red", "yellow", "green")) + 
  theme(axis.text.x = element_blank(), 
        axis.title.x = element_blank()) # to remove the continent names from the x axis.
```

Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population. Just try something other than the plain vanilla mean.
```{r}

```

How is life expectancy changing over time on different continents?
```{r}
data <- gapminder %>% 
  group_by(continent) %>%
  summarize(avgLifeExp = mean(lifeExp)) %>%
  arrange(avgLifeExp) # optional
knitr::kable(data)

# And now also grouping by year
avgLifeExpPerYear <- gapminder %>% 
  group_by(continent, year) %>%
  summarize(avgLifeExp = mean(lifeExp)) %>%
  arrange(year) # optional
knitr::kable(avgLifeExpPerYear)

# Plotting these out directly
ggplot(avgLifeExpPerYear,
       aes(x=year, y=avgLifeExp)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~ continent)

# Similar to above, we can plot a violin plot to understand the trends.
ggplot(gapminder, aes(x=continent, y=lifeExp, fill=continent)) +
  geom_violin() +
  facet_wrap(~ year) +
  scale_fill_manual(values = c("blue", "red", "yellow", "green")) + 
  theme(axis.text.x = element_blank(), 
        axis.title.x = element_blank()) # to remove the continent names from the x axis.
```

Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: Compute some measure of worldwide life expectancy – you decide – a mean or median or some other quantile or perhaps your current age. Then determine how many countries on each continent have a life expectancy less than this benchmark, for each year.
```{r}

```

Find countries with interesting stories. Open-ended and, therefore, hard. Promising but unsuccessful attempts are encouraged. This will generate interesting questions to follow up on in class.
```{r}

```


Some things I found useful for this homework:
[Merging data](https://www.statmethods.net/management/merging.html) 

